---
title: "Cyclistic - Bike Share Case Study"
author: "Asmita Meshram"
date: "27/05/2021"
output:
  html_document: default
  pdf_document: default
---


### A Capstone Project for Google Data Analytics Professional Certification


## Business Task:

Understand how casual riders and annual members of Cyclistic - Bike Share use Cyclistic bikes differently and come up with efficient recommendations to convert casual riders into annual members



## Stakeholders

1. Lily Moreno - Director of Marketing

2. Cyclistic marketing analytics team

3. Cyclistic executive team


## Ask Phase of Data Analysis

As it is clear that we want to convert casual members into annual members, we can come up with these SMART questions -

1. How do annual members and casual riders use Cyclistic bikes differently?

2. Why would casual riders buy Cyclistic annual memberships?

3. How can Cyclistic use digital media to influence casual riders to become members?


## Preparing Our Data

The data set we'll use can be downloaded from [here](https://divvy-tripdata.s3.amazonaws.com/index.html). This dataset is publically available under this [license](https://www.divvybikes.com/data-license-agreement)



##### Installing the required libraries in R

```{r installing libraries}
# libraries are already installed using install.packages()
```


```{r loading installed libraries}
library("tidyverse")
library("lubridate")
library("ggplot2")
```


```{r setting up working directory}
getwd()
```


#### Uploading the datasets

```{r importing datsets}
data_may20 <- read_csv("DataSet/Clean_ds_null_removed/202005-divvy-tripdata.csv")
data_jun20 <- read_csv("DataSet/Clean_ds_null_removed/202006-divvy-tripdata.csv")
data_jul20 <- read_csv("DataSet/Clean_ds_null_removed/202007-divvy-tripdata.csv")
data_aug20 <- read_csv("DataSet/Clean_ds_null_removed/202008-divvy-tripdata.csv")
data_sep20 <- read_csv("DataSet/Clean_ds_null_removed/202009-divvy-tripdata.csv")
data_oct20 <- read_csv("DataSet/Clean_ds_null_removed/202010-divvy-tripdata.csv")
data_nov20 <- read_csv("DataSet/Clean_ds_null_removed/202011-divvy-tripdata.csv")
data_dec20 <- read_csv("DataSet/Clean_ds_null_removed/202012-divvy-tripdata.csv")
data_jan21 <- read_csv("DataSet/Clean_ds_null_removed/202101-divvy-tripdata.csv")
data_feb21 <- read_csv("DataSet/Clean_ds_null_removed/202102-divvy-tripdata.csv")
data_mar21 <- read_csv("DataSet/Clean_ds_null_removed/202103-divvy-tripdata.csv")
data_apr21 <- read_csv("DataSet/Clean_ds_null_removed/202104-divvy-tripdata.csv")
```


## Processing and Cleaning the datasets


#### Inspecting dataframes

```{r }
str(data_may20)
str(data_jun20)
str(data_jul20)
str(data_aug20)
str(data_sep20)
str(data_oct20)
str(data_nov20)
str(data_dec20)
str(data_jan21)
str(data_feb21)
str(data_mar21)
str(data_apr21)
```


We can see in column specifications that start_station_id and end_station_id does not have consistent datatype. So we will change it to character datatype


```{r mutate cols}
data_may20 <-  mutate(data_may20, start_station_id = as.character(start_station_id)
                      ,end_station_id = as.character(end_station_id)) 
data_jun20 <-  mutate(data_jun20, start_station_id = as.character(start_station_id)
                      ,end_station_id = as.character(end_station_id)) 
data_jul20 <-  mutate(data_jul20, start_station_id = as.character(start_station_id)
                      ,end_station_id = as.character(end_station_id)) 
data_aug20 <-  mutate(data_aug20, start_station_id = as.character(start_station_id)
                      ,end_station_id = as.character(end_station_id)) 
data_sep20 <-  mutate(data_sep20, start_station_id = as.character(start_station_id)
                      ,end_station_id = as.character(end_station_id)) 
data_oct20 <-  mutate(data_oct20, start_station_id = as.character(start_station_id)
                      ,end_station_id = as.character(end_station_id))
data_nov20 <-  mutate(data_nov20, start_station_id = as.character(start_station_id)
                      ,end_station_id = as.character(end_station_id)) 
data_dec20 <-  mutate(data_dec20, start_station_id = as.character(start_station_id)
                      ,end_station_id = as.character(end_station_id)) 
data_jan21 <-  mutate(data_jan21, start_station_id = as.character(start_station_id)
                      ,end_station_id = as.character(end_station_id))
data_feb21 <-  mutate(data_feb21, start_station_id = as.character(start_station_id)
                      ,end_station_id = as.character(end_station_id)) 
data_mar21 <-  mutate(data_mar21, start_station_id = as.character(start_station_id)
                      ,end_station_id = as.character(end_station_id)) 
data_apr21 <-  mutate(data_apr21, start_station_id = as.character(start_station_id)
                      ,end_station_id = as.character(end_station_id))
```


#### Combining all datasets into a single dataset

```{r binding datasets}
trips <- bind_rows(data_may20,data_jun20,data_jul20,data_aug20,data_sep20,data_oct20,data_nov20,data_dec20,data_jan21,data_feb21,data_mar21,data_apr21)
```


Longitude and latitude columns can be dropped as there is no need of them

```{r dropping unwanted cols}

trips <- trips %>%
  select(-c(start_lat,start_lng,end_lat,end_lng))
```


### Formatting data for our convenience

```{r adding date, month and year col}
trips$date <- as.Date(trips$started_at, "%d-%m-%Y")

trips$month <- format(as.Date(trips$date),"%m")

trips$day <- format(as.Date(trips$date),"%d")

trips$year <- format(as.Date(trips$date), "%Y")

trips$month_group <- format(as.Date(trips$date), "%Y-%m")
```


#### Summarizing data for analysis

```{r summarise data}
colnames(trips)

nrow(trips)

dim(trips)

head(trips)

str(trips)

summary(trips)
```


The ride_length cannot be summarized as it is in HH:MM:SS format. so let's convert it to numeric representing seconds only

```{r ride length in sec}
trips$ride_length <- hms(trips$ride_length)
trips$ride_length <- as.numeric(trips$ride_length)
```

```{r renaming ride_length}

(trips <- rename(trips,
                 ride_length_in_sec = ride_length))
```




## Analyse Phase


```{r summarizing}
mean(trips$ride_length_in_sec)

median(trips$ride_length_in_sec)

max(trips$ride_length_in_sec)

min(trips$ride_length_in_sec)

```


```{r aggregate}
aggregate(trips$ride_length_in_sec ~ trips$member_casual, FUN = mean)

aggregate(trips$ride_length_in_sec ~ trips$member_casual, FUN = median)

aggregate(trips$ride_length_in_sec ~ trips$member_casual, FUN = max)

aggregate(trips$ride_length_in_sec ~ trips$member_casual, FUN = min)


aggregate(trips$ride_length_in_sec ~ trips$member_casual + trips$day_of_week, FUN = mean)

```


```{r quick summary}


trips %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual,weekday) %>%
  summarise(number_of_rides = n(), average_duration = mean(ride_length_in_sec)) %>%
  arrange(member_casual, weekday)


trips %>%
  group_by(month_group, member_casual) %>%
  summarise(number_of_rides = n(), average_duration = mean(ride_length_in_sec)) %>%
  arrange(month_group, member_casual)
```



## Plotting our data

```{r plot}
trips %>%
  group_by(member_casual, day_of_week) %>%
  summarise(number_of_rides = n(), average_duration = mean(ride_length_in_sec)) %>%
  arrange(member_casual, day_of_week) %>%
  ggplot(aes(x = day_of_week, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge", width = 0.8)
```

```{r plot2}
trips %>%
  group_by(month_group, member_casual) %>%
  summarise(number_of_rides = n(), average_duration = mean(ride_length_in_sec)) %>%
  arrange(month_group, member_casual) %>%
  ggplot(aes(x = month_group, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge", width = 0.5) +
  theme(axis.text.x = element_text(angle = 90)) +
  ggtitle("Average duration of rides from May 2020 to Apr 2021")
```

```{r plot3}
trips %>%
  group_by(rideable_type, member_casual) %>%
  summarise(number_of_rides = n(), average_duration = mean(ride_length_in_sec)) %>%
  arrange(rideable_type, member_casual) %>%
  ggplot(aes(x = rideable_type, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge", width = 0.5) +
  theme(axis.text.x = element_text(angle = 90)) +
  ggtitle("Average duration of rides for each ride type")
```



##### Export data for further Analysis

```{r}
agg_data = aggregate(trips$ride_length_in_sec ~ trips$member_casual + trips$day_of_week, FUN = mean)
# write.csv(agg_data, file = 'D:\\Main Folder\\Google Data Analytics Course\\CaseStudy\\Cyclistic_Case_Study\\DataSet\\agg_data.csv')
# write.csv(trips, file = 'D:\\Main Folder\\Google Data Analytics Course\\Case Study\\Cyclistic_Case_Study\\DataSet\\trips.csv')
```
